GVN-Hoist: Hoisting Computations from Branches

### 2016-08-08 hoist.tex
legality

The process of computing GVN can be on-demand (as we come across an instruction)
or, precomputed (computing GVN of all the instructions beforehand). Which
process to choose is determined by the scope of code-hoisting we want to
perform. A pessimistic approach where we want to hoist limited set of
instructions from the sibing branches as we iterate the DFS tree bottom-up, it
is sufficient to compute values on-demand. Whereas, in the optimistic approach,
where we want to hoist as many instructions as possible, would require values to
be precomputed.

@@

\subsection{Checking for legality}
Finding a hoisting point i.e., a common dominator of the set of instructions. If
such a dominator is found then checking if all the use-operands of the set of
instructions are available or not. In some cases if the operands are not
available, it is possible to reinstantiate (remateralize) the use-operands, thus
making the passing the legality check. Checking that the side-effects of the
computations does not intersect with any side-effect between the instructions to
be hoisted and their hoisting point. For memory operations like loads, stores,
calls etc., it is also required to check that all the paths from the hoisting
point to the end of the function should have the exact memory operation
available, in order to guarantee correctness.

@@

\subsubsection{Hoisting loads}
At the hoisting point the availability of operands to the load (an address) is
checked. If it is not available we try to compute the addresss if that is
possible.  Along the path, from current position of the load instruction through
the hoisting point, we find if there are aliasing stores, we find if any of the
basic blocks are landing pads or targets of indirect branches. In those cases we
discard the candidate.

\subsubsection{Hoisting stores}
We need to check the dependency requirements similar to the hoisting of
loads. We need to check that the operands of the store instruction are available
at the hoisting point, that there are no aliasing loads or store along the path
from the current position through the hoisting point and, none of the basic
blocks along the path have their address taken.

\subsubsection{Hoisting calls}
We have to check that the loads/stores in the call does not create hazard with
any loads/stores (in the caller) in the path from callsite through the hoisting
point, just like we did for loads and stores.

@@

\subsubsection{Hoisting scalars}
Care must be taken in case of hoisting scalars too far, as that may increase
register pressure and result in spills. For example hoisting a scalar past a
call (Section-\ref{cost:across-calls}). In our current implementation we hoist
scalars past a call only when optimizing for code-side (-Os). Another way to
mitigate this problem is be to reinstantiate (rematerialize) the computation
after a call (may be as a different pass).

@@

\subsection{Hoisting bottom up with optimistic approach}
Optimistic means collect the set of all potential redundancies and
then discard the bad ones.

@@

\subsubsection{Pessimistic approach}
To extend the triangle approach to the entire function, the algorithm traverses
the basic blocks in the invers depth-first order. All the leaf nodes are visited
before the non-leaf nodes because instructions are hoisted upwards. We start
from a branch and collect the GVN of all the instructions. Then in the sibling
basic block we find if there are any identical computations. Once we find an
identical computation we try to hoist it to the nearest common dominator which
is the immediate predecessor of both the basic blocks. We traverse the basic
block from top to down so that the once an instruction is hoisted it creates
opportunities for other dependent instructions. After all the sibling basic
blocks have all the identical computations hoisted, we continue the hoisting
process with their parent and its sibling of their parent until we reach the
beginning of the function.

>>>>

The process of computing GVN can be on-demand (as we come across an instruction)
or, precomputed (computing GVN of all the instructions beforehand). Which
process to choose is determined by the scope of code-hoisting we want to
perform. In a pessimistic approach (Section-\ref{subsec:pessimistic}) where we
want to hoist limited set of instructions from the sibing branches as we iterate
the DFS tree bottom-up, it is sufficient to compute values on-demand. Whereas,
in the optimistic approach (Section-\ref{subsec:optimistic}), where we want to
hoist as many instructions as possible, would require values to be precomputed.

@@

\subsection{Checking for legality}
Finding a hoisting point i.e., a common dominator of the set of instructions. If
such a dominator is found then checking if all the use-operands of the set of
instructions are available or not. In some cases if the operands are not
available, it is possible to reinstantiate (remateralize) the use-operands, thus
passing the legality check. Checking that the side-effects of the computations
does not intersect with any side-effects between the instructions to be hoisted
and their hoisting point. For memory operations like loads, stores, calls etc.,
it is also required to check that all the paths from the hoisting point to the
end of the function should execute the exact instruction, in order to guarantee
correctness. Moreover, hoisting of memory operations should not be done on paths
which have indirect branch targets e.g., landing pad, case statements, goto
labels etc., because it becomes tricky to prove that all the paths from hoisting
point to the end of the function would execute the instruction.

@@

\subsubsection{Hoisting loads}
At the hoisting point the availability of operands to the load (an address) is
checked. If it is not available we try to compute the addresss if that is
possible.  Along the path, from current position of the load instruction through
the hoisting point, we find if there are aliasing stores. In those cases we
discard the candidate.

\subsubsection{Hoisting stores}
We need to check the dependency requirements similar to the hoisting of
loads. We need to check that the operands of the store instruction are available
at the hoisting point, that there are no aliasing loads or store along the path
from the current position through the hoisting point.

\subsubsection{Hoisting calls}
For hoisting the call instructions, we check that the loads/stores in the call
does not create hazard with any loads/stores (in the caller) in the path from
callsite through the hoisting point, just like we did for loads and stores.

@@

\subsubsection{Hoisting scalars}
Care must be taken in case of hoisting scalars too far, as that may increase
register pressure and result in spills. For example hoisting a scalar past a
call (Section-\ref{cost:across-calls}). In our current implementation we hoist
scalars past a call only when optimizing for code-side (-Os). Another way to
mitigate this problem is be to reinstantiate (rematerialize) the computation
after a call (may be as a different optimization pass).

@@

\subsection{Hoisting bottom up with optimistic approach}
\label{subsec:optimistic}
Optimistic means collect the set of all potential redundancies and
then discard the bad ones.

@@

\subsubsection{Pessimistic approach}
\label{subsec:pessimistic}
To extend the triangle approach to the entire function, the algorithm traverses
the basic blocks in the invers depth-first order. All the leaf nodes are visited
before the non-leaf nodes because instructions are hoisted upwards. We start
from a branch and collect the GVN of all the instructions. Then in the sibling
basic block we find if there are any identical computations. Once we find an
identical computation we try to hoist it to the nearest common dominator which
is the immediate predecessor of both the basic blocks. We traverse the basic
block from top to down so that the once an instruction is hoisted it creates
opportunities for other dependent instructions. After all the sibling basic
blocks have all the identical computations hoisted, we continue the hoisting
process with their parent and its sibling of their parent until we reach the
beginning of the function.
